---
title: "HW2"
author: "Sukhman Parhar"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set

library(tidyverse)
library(ISLR)
library(glmnet)
library(caret)
library(tidymodels)
library(corrplot)
library(ggplot2)
library(plotmo)
library(ggrepel)
library(pls)
library(pdp)
library(earth)
library(splines)
library(mgcv)
```

## load and clean the data

```{r}
# Load in the data 
college <- read_csv("College.csv")

# look at the data
#str(college)
#dim(college)
#summary(college)
#colSums(is.na(college))

# clean the data

# remove college var
college$College <- NULL

# cap grad rate at 100% anything above is impossible
college <- college[college$Grad.Rate <= 100, ]
```

## Data partition

```{r}
set.seed(2)

data_split <- initial_split(college, prop = 0.8)

training_data <- training(data_split)
testing_data  <- testing(data_split)
```

## Question 1: Smoothing Spline Models 

```{r}
# predictor and response
x <- training_data$Top10perc
y <- training_data$Outstate

# smoothing splines with different df
fit3  <- smooth.spline(x, y, df = 3)
fit5  <- smooth.spline(x, y, df = 5)
fit7  <- smooth.spline(x, y, df = 7)
fit10 <- smooth.spline(x, y, df = 10)
```

### Plot the lines

```{r}
plot(x, y,
     col = "grey",
     pch = 16,
     xlab = "Top10perc",
     ylab = "Outstate")

lines(predict(fit3), col = "blue", lwd = 2)
lines(predict(fit5), col = "green", lwd = 2)
lines(predict(fit7), col = "orange", lwd = 2)
lines(predict(fit10), col = "red", lwd = 2)

legend("topleft",
       legend = c("df=3","df=5","df=7","df=10"),
       col = c("blue","green","orange","red"),
       lwd = 2)
```

### Find optimal fit 

```{r}
fit.opt <- smooth.spline(training_data$Top10perc,
                         training_data$Outstate)

fit.opt$df

plot(training_data$Top10perc,
     training_data$Outstate,
     col = "grey",
     pch = 16,
     xlab = "Top10perc",
     ylab = "Outstate")

lines(fit.opt, col = "black", lwd = 3)
```

##
When we increase the df the smoothing spline becomes more flexible and follows the data more. Meaning that the lines we see for lower df are smoother and higher df curve more. 

The optimal degree of freedom was selected using generalized cross-validation through the smooth.spline() function. The appropriate df was 2.29 which means there is a smooth relationship between Top10perc and Outstate tuition.

## Question 2: MARS

```{r}
# matrix of predictors
x <- model.matrix(Outstate ~ ., training_data)[, -1]

# vector of response
y <- training_data$Outstate

# ten-fold CV
ctrl1 <- trainControl(method = "cv", number = 10)

# tuning grid
mars_grid <- expand.grid(degree = 1:3,
                         nprune = 2:15)

set.seed(2)
mars.fit <- train(x, y,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl1)

ggplot(mars.fit)

mars.fit$bestTune
coef(mars.fit$finalModel)
```

### Regression Function 
$$
\hat{Outstate} =
8935.38
-0.756\,h(Expend-15736)
-0.703\,h(5720-Room.Board)
-1.665\,h(1365-F.Undergrad)
-90.907\,h(21-perc.alumni)
+0.423\,h(Apps-2652)
-1.941\,h(Enroll-958)
+6.206\,h(958-Enroll)
-2.095\,h(2059-Accept)
+22.080\,h(Grad.Rate-42)
+0.735\,h(Expend-6881)
+0.759\,h(1300-Personal)
$$

### PDP

```{r}
p1 <- pdp::partial(mars.fit, pred.var = "Room.Board", grid.resolution = 10) |> autoplot()

p1
```

### Test MSE (COME BACK TO THIS )